---
title: "BIOSTAT 724 - Homework 1"
author: "Austin Allen"
---

## Problem 1

In answering a question on a multiple-choice test,
a student either knows the answer or guesses. Let p be the
probability that the student knows the answer and 1- p the
probability that the student guesses. Assume that a student
who guesses at the answer will be correct with probability
1/m, where m is the number of multiple-choice alternatives.
What is the conditional probability that a student knew the
answer to a question, given that he or she answered it
correctly? What happens when m is very large? What
happens when p is close to 0?

> Let's solve this using Baye's Theorem. First, it'll be helpful to define some terms:
>
> - $P(\text{Knows}) = p$
> - $P(\text{Guess}) - 1 - p$
> - $P(\text{Correct | Guess}) = \frac{1}{m}$
> - We want to know $P(\text{Knows | Correct})$


\begin{align*}
P(\text{Knows | Correct}) &= \frac{P(\text{Correct | Knows})P(\text{Knows})}{P(\text{Correct | Knows})P(\text{Knows}) + P(\text{Correct | Guess})P(\text{Guess})}\\
&= \frac{(1)(p)}{(1)(p) + (\frac{1}{m})(1 - p)}\\
&= \frac{p}{p + \frac{1 - p}{m}}\\
\end{align*}

> We can see from this result that as m gets arbitrarily large, the denominator approaches p and the probability goes to 1. This aligns with our intuition because a correct guess among a larger number of choices would be less likely. Likewise, if p is close to 0, the probability is also close to zero.

## Problem 2

Suppose urn H is filled with 40% green balls and
60% red balls, and urn T is filled with 60% green balls and
40% red balls. Someone will flip a coin and then select a
ball from urn H or urn T depending on whether the coin
lands heads or tails, respectively. Let X be 1 or 0 if the coin
lands heads or tails, and let Y be 1 or 0 if the ball is green
or red.



### Part a

Write out the joint distribution of X and Y in a table.

```{python}
# | echo: false
import pandas as pd

# Define the joint probability table
df = pd.DataFrame(
    {0: [0.2, 0.3], 1: [0.3, 0.2]},
    index=pd.Index([0, 1], name="Y"),
    columns=pd.Index([0, 1], name="X"),
)

# Add marginal probabilities
df["P(Y)"] = df.sum(axis=1)  # Row sums (marginal P(Y))
df.loc["P(X)"] = df.sum(axis=0)  # Column sums (marginal P(X))

# Display the table
print(df)

```

### Part b

Find E(Y). What is the probability that the ball is
green?

$E(Y) = \sum y\cdot f_Y(y)$. In our table, we see that Y can take on two values, 0 and 1. We can use the marginal probability to find that  $\sum y\cdot f_Y(y) = (0)(0.5) + (1)(0.5) = 0.5$. Thus, $P(\text{Ball is green}) = 0.5$.

### Part c 

Find Var(Y|X=0), Var(Y|X=1) and Var(Y). Thinking
of variance as measuring uncertainty, explain
intuitively why one of these variances is larger than
the others.

> Intuitively, I would expect the variance of Y without any additional information regarding X to be greater than either variances of Y given X = 0 and X = 1. That is, I expect $max(Var(Y), Var(Y|X=0), Var(Y|X=1)) = Var(Y)$. My reasoning is that knowing something about X reduces our uncertainty of Y. Let's see if I'm correct.

$$
\begin{align*}
Var(Y) &= \sum_{i = 1}^n (Y_i - E[Y])^2\cdot f(Y)\\
&= (0 -0.5)^2(0.5) + (1-0.5)^2(0.5)\\
&= 0.25\\
Var(Y|X=x) &= \sum_{i =1}^n (Y_i - E[Y | x])^2\cdot f(y|x)\\
E(Y|X=x) &= \sum_{i = 1}^nY_i\cdot f(y|x)\\
E(Y|X=0) &= (0)(0.4) + (1)(0.6)\\
&= 0.6\\
E(Y|X=1) &= (0)(0.6) + (1)(0.4)\\
&= 0.4\\
Var(Y|X=0) &= (0 - 0.6)^2(0.4) + (1 - 0.6)^2(0.6)\\
&= 0.24\\
Var(Y|X=1) &= (0 - 0.4)^2(0.6) + (1 - 0.4)^2(0.4)\\
&= 0.24\\
\end{align*}
$$

> As I guessed, knowing something about X does reduce the uncertainty around Y. We see that the conditional variances are both less than the variance of Y, even if it is by a small amount. 


### Part d 

Suppose you see that the ball is green. What is the
probability that the coin turned up tails?

> Let's solve this in terms of X and Y. Thus, we want $P(X=0|Y=1)$, which is easily solved using Bayes' Theorem:


\begin{align*}
P(X=0|Y=1) &= \frac{P(Y=1|X=0)P(X=0)}{P(Y=1|X=0)P(X=0) + P(Y=1|X=1)P(X=1)}\\
&= \frac{(0.6)(0.5)}{(0.6)(0.5) + (0.4)(0.5)}\\
&= 0.6
\end{align*}



## Problem 3

Let $\theta$ be a Beta(a,b) random variable, and suppose
that the conditional distribution of X, given that $\theta$, is
Binomial(n, $\theta$). Find the (marginal) probability mass
function of X.

> We can solve this by integrating out $\theta$ from the joint distribution. 

\begin{align*}
f_X(X) &= \int P(X=x|\theta)P(\theta)d\theta\\
&= \int {n \choose x} \theta^x(1 - \theta)^{n - x}\theta^{a-1}(1-\theta)^{b-1}\frac{\Gamma(a + \beta)}{\Gamma(a)\Gamma(b)}d\theta\\
&= \int {n \choose x} \theta^{x+a-1}(1 - \theta)^{n + b- x-1}\frac{\Gamma(a + \beta)}{\Gamma(a)\Gamma(b)}d\theta\\
&= {n \choose x}\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)} \int\theta^{x+a-1}(1 - \theta)^{n + b- x-1}d\theta\\
&= {n \choose x}\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)}\frac{\Gamma(x + a - 1)\Gamma(b + n - x - 1)}{\Gamma(a + b + n - 2)}\int \theta^{x+a-1}(1 - \theta)^{n + b- x-1}d\theta \frac{\Gamma(a + b + n - 2)}{\Gamma(x + a - 1)\Gamma(b + n - x - 1)}d\theta\\
&= {n \choose x}\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)}\frac{\Gamma(x + a - 1)\Gamma(b + n - x - 1)}{\Gamma(a + b + n - 2)} (1)\\
\end{align*}



## Problem 4

Full conditionals: Let X,Y,Z be random variables
with joint density (discrete or continuous) p(x, y, z) ∝
f(x, z)g(y, z)h(z). Show that

(a) p(x|y, z) ∝ f(x, z), i.e. , p(x|y, z) is a function of x
and z;
(b) p(y|x, z) \propto g(y, z), i.e., p(y|x, z) is a function of y and z;
(c) X and Y are conditionally independent given Z.

### Part a 


\begin{align*}
p(x|y, z) &= \frac{p(x, y, z)}{p(y)p(z)}\\
\implies p(x|y, z) &\propto p(x, y, z)\\
\implies p(x|y, z) &\propto f(x, z)g(y, z)h(z)\\
\end{align*}

> Note that the only part of this that depends on X is $f(x, z)$, which implies that $p(x|y, z) \propto f(x, z)$. 

### Part b 

> Let's solve this in the exact same way. 


\begin{align*}
p(y|x, z) &= \frac{p(x, y, z)}{p(x)p(z)}\\
\implies p(y|x, z) &\propto p(x, y, z)\\
\implies p(y|x, z) &\propto f(x, z)g(y, z)h(z)\\
\end{align*}

> Similarly, the only portion of $f(x, z)g(y, z)h(z)$ that depends on y is the term $g(y, z)$. Thus, $p(y|x, z) \propto g(y, z)$. 

### Part c 

> To be conditionally independent, we need to show that $p(x, y|z) \propto p(x|z)p(y|z)$. 
> 
> Let's begin with the same approach that we followed in parts a and b. We'll obviously need to get a little fancier for this part, but let's first show that $p(x, y|z) \propto f(x,z)g(y,z):

\begin{align*}
p(y, x| z) &= \frac{p(x, y, z)}{p(z)}\\
\implies p(y,x| z) &\propto p(x, y, z)\\
\implies p(y,x| z) &\propto f(x, z)g(y, z)h(z)\\
\implies p(y,x| z) &\propto f(x, z)g(y, z)\\
\end{align*}

> Next, we need to think about $p(x|z)$ and $p(y|z)$. 


\begin{align*}
p(x|z) &= \frac{p(x,z)}{p(z)}\\
p(x,z) \propto f(x,z)f(z)\\
\implies p(x|z)\propto f(x,z)\\
p(y|z) &= \frac{p(y,z)}{p(z)}\\
p(y,z) \propto g(y,z)f(z)\\
\implies p(y|z)\propto g(y,z)\\
\end{align*}

> Combining these results, we see that $p(x, y | z)\propto f(x,z)g(y,z)\implies p(x, y |z) \propto p(x|z)p(y|z)$. Thus, x and y are conditionally independent. 


## Part 5

PH 3.1


### Part a

Assume Y1, . . . , Y100 are, conditional on θ, i.i.d. binary random variables with expectation θ. Write down the joint distribution of Pr(Y1 = y1, . . . , Y100 = y100|θ) in a compact form. Also write down the form of $Pr(\sum PYi = y|θ)$.


\begin{align*}
f(y_1, ..., y_{100}|\theta) &= \prod_{i=1}^{100}f(y_i|\theta)\\
&= \theta^{\sum_{i = 1}^{100}y_i}(1 - \theta)^{n - \sum_{i = 1}^{100}y_i}\\
P\left(\sum_{i = 1}^{100}Y = y_i|\theta\right) &= {n \choose \sum y_i} \theta^{\sum y_i}(1 - \theta)^{n - \sum y_i}
\end{align*}



### Part b 

For the moment, suppose you believed that θ ∈ {0.0, 0.1, . . . , 0.9, 1.0}. Given that the results of the survey were $\sum Yi = 57$, compute $Pr(\sum Yi = 57|θ)$ for each of these 11 values of θ and plot these probabilities as a function of θ.


```{r warning=FALSE, message=FALSE}
###################################### R ######################################
library(ggplot2)

# Define a function that will spit out the probability we're interested in
# computing
p_y_theta <- function(theta, n, y) {
    # y: a sum of bernouli trials
    # theta: a vector of values between 0 and 1 (can be length 1)
    # n: number of bernouli trials
    return(choose(n, y) * theta**y * (1 - theta)**(n - y))
}

# Define our primary input values (theta) and other given values (n, y)
theta <- seq(0, 1, length.out = 11)
n <- 100
y <- 57

# Figure out where the highest probability lies
max_prob_theta <- theta[
    which(p_y_theta(theta, n, y) == max(p_y_theta(theta, n, y)))
]
# Plot the outputs as nice and pretty as possible
plot_data <- data.frame(theta = theta, y = p_y_theta(theta, n, y))
ggplot(plot_data, aes(x = theta, y = y)) +
    geom_point(col = "skyblue", size = 2) +
    geom_line(col = "skyblue", lwd = 0.5, lty = 2) +
    geom_vline(xintercept = max_prob_theta, col = "firebrick", lwd = 1, lty = 2) +
    labs(title = "Pr(sum(Y) = 57 | theta)", x = "Theta", y = "Probability") +
    theme_bw() +
    theme(
        text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
###############################################################################
```

> The plot shows a maximum at `r max_prob_theta`. 

### Part c 

Now suppose you originally had no prior information to believe one of these θ-values over another, and so Pr(θ = 0.0) = Pr(θ = 0.1) = · · · = Pr(θ = 0.9) = Pr(θ = 1.0). Use Bayes’ rule to compute $p(θ| \sum Yi = 57)$ for each θ-value. Make a plot of this posterior distribution as a function of θ.

> I want to note that I used generative AI to assist me with the following code for this problem. I derived the the math to implement, and ChatGPT assisted with the coding: 

```{r}
###################################### R ######################################

# Define the function to compute posterior probability using Bayes' theorem
p_theta_bayes <- function(theta, n, y) {
    # Compute P(Y = 57 | θ) for each θ
    likelihood <- choose(n, y) * theta^y * (1 - theta)^(n - y)

    # Prior probabilities (uniform over 11 values)
    prior <- rep(1 / length(theta), length(theta))

    # Compute P(Y = 57) as the sum of likelihood * prior over all θ
    p_y <- sum(likelihood * prior)

    # Compute posterior using Bayes' rule
    posterior <- (likelihood * prior) / p_y

    return(posterior)
}

# Define theta values and known values of n and y
theta <- seq(0, 1, length.out = 11)
n <- 100
y <- 57

# Compute the posterior probabilities
posterior_probs <- p_theta_bayes(theta, n, y)

# Plot the posterior distribution
plot_data <- data.frame(theta = theta, posterior = posterior_probs)

ggplot(plot_data, aes(x = theta, y = posterior)) +
    geom_point(col = "blue", size = 2) +
    geom_line(col = "blue", lwd = 0.5, lty = 2) +
    labs(
        title = "Posterior Probability Distribution P(θ | Y = 57)",
        x = "Theta", y = "Posterior Probability"
    ) +
    theme_bw() +
    theme(
        text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
###############################################################################
```


### Part d 

Now suppose you allow θ to be any value in the interval [0, 1]. Using
the uniform prior density for θ, so that p(θ) = 1, plot the posterior
density $p(θ) × Pr(\sum Yi = 57|θ)$ as a function of θ.

```{r}
###################################### R ######################################
library(ggplot2)

# Define function to compute posterior density
posterior_density <- function(theta_vals, n, y) {
    # Compute likelihood P(Y = 57 | θ)
    likelihood <- choose(n, y) * theta_vals^y * (1 - theta_vals)^(n - y)

    # Since prior is uniform (1), posterior is proportional to likelihood
    posterior <- likelihood

    return(posterior)
}

# Define a continuous range of theta values
theta_vals <- seq(0, 1, length.out = 1000) # More points for smooth curve
n <- 100
y <- 57

# Compute posterior density
posterior_vals <- posterior_density(theta_vals, n, y)

# Normalize for better visualization (optional, but makes y-axis interpretable)
posterior_vals <- posterior_vals / max(posterior_vals)

# Plot the posterior density
plot_data <- data.frame(theta = theta_vals, posterior = posterior_vals)

ggplot(plot_data, aes(x = theta, y = posterior)) +
    geom_line(col = "skyblue", lwd = 1) +
    labs(
        title = "Posterior Density p(θ | Y = 57)",
        x = "Theta", y = "Density (unnormalized)"
    ) +
    theme_bw() +
    theme(
        text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
###############################################################################

```

### Part e 


```{r}
###################################### R ######################################

# Define a function to return a posterior Beta(1+57, 1+100-57) density given
# an input of theta values
posterior_beta_density <- function(theta_vals) {
    return(dbeta(theta_vals, 1 + 57, 1 + 100 - 57))
}

# Get a set of density values like we did in part d
posterior_probs <- posterior_beta_density(theta_vals)

# Combine into a dataframe
plot_data <- data.frame(theta = theta_vals, density = posterior_probs)

# Now let's get back to plotting!
ggplot(plot_data, aes(x = theta, y = density)) +
    geom_line(col = "skyblue", lwd = 1) +
    labs(
        title = "Posterior Density Beta(1+57, 100+1-57)",
        x = "Theta", y = "Density"
    ) +
    theme_bw() +
    theme(
        text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
###############################################################################
```




## Part 6 



PH 3.2