---
title: "BIOSTAT 724 - Homework 1"
author: "Austin Allen"
output:
  pdf_document:
    latex_engine: xelatex
---

## Problem 1

In answering a question on a multiple-choice test,
a student either knows the answer or guesses. Let p be the
probability that the student knows the answer and 1- p the
probability that the student guesses. Assume that a student
who guesses at the answer will be correct with probability
1/m, where m is the number of multiple-choice alternatives.
What is the conditional probability that a student knew the
answer to a question, given that he or she answered it
correctly? What happens when m is very large? What
happens when p is close to 0?

> Let's solve this using Baye's Theorem. First, it'll be helpful to define some terms:
>
> - $P(\text{Knows}) = p$
> - $P(\text{Guess}) - 1 - p$
> - $P(\text{Correct | Guess}) = \frac{1}{m}$
> - We want to know $P(\text{Knows | Correct})$


\begin{align*}
P(\text{Knows | Correct}) &= \frac{P(\text{Correct | Knows})P(\text{Knows})}{P(\text{Correct | Knows})P(\text{Knows}) + P(\text{Correct | Guess})P(\text{Guess})}\\
&= \frac{(1)(p)}{(1)(p) + (\frac{1}{m})(1 - p)}\\
&= \frac{p}{p + \frac{1 - p}{m}}\\
\end{align*}

> We can see from this result that as m gets arbitrarily large, the denominator approaches p and the probability goes to 1. This aligns with our intuition because a correct guess among a larger number of choices would be less likely. Likewise, if p is close to 0, the probability is also close to zero.

## Problem 2

Suppose urn H is filled with 40% green balls and
60% red balls, and urn T is filled with 60% green balls and
40% red balls. Someone will flip a coin and then select a
ball from urn H or urn T depending on whether the coin
lands heads or tails, respectively. Let X be 1 or 0 if the coin
lands heads or tails, and let Y be 1 or 0 if the ball is green
or red.



### Part a

Write out the joint distribution of X and Y in a table.

```{r echo = FALSE}

# Define the joint probability table
df <- data.frame(
    X0 = c(0.2, 0.3), X1 = c(0.3, 0.2)
)

# Add rownames for Y 
row.names(df) <- c("Y0", "Y1")

# Display the table
pander::pander(df)

```

### Part b

Find E(Y). What is the probability that the ball is
green?

$E(Y) = \sum y\cdot f_Y(y)$. In our table, we see that Y can take on two values, 0 and 1. We can use the marginal probability to find that  $\sum y\cdot f_Y(y) = (0)(0.5) + (1)(0.5) = 0.5$. Thus, $P(\text{Ball is green}) = 0.5$.

### Part c 

Find Var(Y|X=0), Var(Y|X=1) and Var(Y). Thinking
of variance as measuring uncertainty, explain
intuitively why one of these variances is larger than
the others.

> Intuitively, I would expect the variance of Y without any additional information regarding X to be greater than either variances of Y given X = 0 and X = 1. That is, I expect $max(Var(Y), Var(Y|X=0), Var(Y|X=1)) = Var(Y)$. My reasoning is that knowing something about X reduces our uncertainty of Y. Let's see if I'm correct.


\begin{align*}
Var(Y) &= \sum_{i = 1}^n (Y_i - E[Y])^2\cdot f(Y)\\
&= (0 -0.5)^2(0.5) + (1-0.5)^2(0.5)\\
&= 0.25\\
Var(Y|X=x) &= \sum_{i =1}^n (Y_i - E[Y | x])^2\cdot f(y|x)\\
E(Y|X=x) &= \sum_{i = 1}^nY_i\cdot f(y|x)\\
E(Y|X=0) &= (0)(0.4) + (1)(0.6)\\
&= 0.6\\
E(Y|X=1) &= (0)(0.6) + (1)(0.4)\\
&= 0.4\\
Var(Y|X=0) &= (0 - 0.6)^2(0.4) + (1 - 0.6)^2(0.6)\\
&= 0.24\\
Var(Y|X=1) &= (0 - 0.4)^2(0.6) + (1 - 0.4)^2(0.4)\\
&= 0.24\\
\end{align*}


> As I guessed, knowing something about X does reduce the uncertainty around Y. We see that the conditional variances are both less than the variance of Y, even if it is by a small amount. 


### Part d 

Suppose you see that the ball is green. What is the
probability that the coin turned up tails?

> Let's solve this in terms of X and Y. Thus, we want $P(X=0|Y=1)$, which is easily solved using Bayes' Theorem:


\begin{align*}
P(X=0|Y=1) &= \frac{P(Y=1|X=0)P(X=0)}{P(Y=1|X=0)P(X=0) + P(Y=1|X=1)P(X=1)}\\
&= \frac{(0.6)(0.5)}{(0.6)(0.5) + (0.4)(0.5)}\\
&= 0.6
\end{align*}



## Problem 3

Let $\theta$ be a Beta(a,b) random variable, and suppose
that the conditional distribution of X, given that $\theta$, is
Binomial(n, $\theta$). Find the (marginal) probability mass
function of X.

> We can solve this by integrating out $\theta$ from the joint distribution. 

\begin{align*}
f_X(X) &= \int P(X=x|\theta)P(\theta)d\theta\\
&= \int {n \choose x} \theta^x(1 - \theta)^{n - x}\theta^{a-1}(1-\theta)^{b-1}\frac{\Gamma(a + \beta)}{\Gamma(a)\Gamma(b)}d\theta\\
&= \int {n \choose x} \theta^{x+a-1}(1 - \theta)^{n + b- x-1}\frac{\Gamma(a + \beta)}{\Gamma(a)\Gamma(b)}d\theta\\
&= {n \choose x}\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)} \int\theta^{x+a-1}(1 - \theta)^{n + b- x-1}d\theta\\
&= {n \choose x}\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)}\frac{\Gamma(x + a - 1)\Gamma(b + n - x - 1)}{\Gamma(a + b + n - 2)}\int \theta^{x+a-1}(1 - \theta)^{n + b- x-1}d\theta \frac{\Gamma(a + b + n - 2)}{\Gamma(x + a - 1)\Gamma(b + n - x - 1)}d\theta\\
&= {n \choose x}\frac{\Gamma(a+b)}{\Gamma(a)\Gamma(b)}\frac{\Gamma(x + a - 1)\Gamma(b + n - x - 1)}{\Gamma(a + b + n - 2)} (1)\\
\end{align*}



## Problem 4

Full conditionals: Let X,Y,Z be random variables
with joint density (discrete or continuous) p(x, y, z) ∝
f(x, z)g(y, z)h(z). Show that

(a) p(x|y, z) ∝ f(x, z), i.e. , p(x|y, z) is a function of x
and z;
(b) p(y|x, z) $\propto$ g(y, z), i.e., p(y|x, z) is a function of y and z;
(c) X and Y are conditionally independent given Z.

### Part a 


\begin{align*}
p(x|y, z) &= \frac{p(x, y, z)}{p(y)p(z)}\\
\implies p(x|y, z) &\propto p(x, y, z)\\
\implies p(x|y, z) &\propto f(x, z)g(y, z)h(z)\\
\end{align*}

> Note that the only part of this that depends on X is $f(x, z)$, which implies that $p(x|y, z) \propto f(x, z)$. 

### Part b 

> Let's solve this in the exact same way. 


\begin{align*}
p(y|x, z) &= \frac{p(x, y, z)}{p(x)p(z)}\\
\implies p(y|x, z) &\propto p(x, y, z)\\
\implies p(y|x, z) &\propto f(x, z)g(y, z)h(z)\\
\end{align*}

> Similarly, the only portion of $f(x, z)g(y, z)h(z)$ that depends on y is the term $g(y, z)$. Thus, $p(y|x, z) \propto g(y, z)$. 

### Part c 

> To be conditionally independent, we need to show that $p(x, y|z) \propto p(x|z)p(y|z)$. 
> 
> Let's begin with the same approach that we followed in parts a and b. We'll obviously need to get a little fancier for this part, but let's first show that $p(x, y|z) \propto f(x,z)g(y,z):

\begin{align*}
p(y, x| z) &= \frac{p(x, y, z)}{p(z)}\\
\implies p(y,x| z) &\propto p(x, y, z)\\
\implies p(y,x| z) &\propto f(x, z)g(y, z)h(z)\\
\implies p(y,x| z) &\propto f(x, z)g(y, z)\\
\end{align*}

> Next, we need to think about $p(x|z)$ and $p(y|z)$. 


\begin{align*}
p(x|z) &= \frac{p(x,z)}{p(z)}\\
p(x,z) \propto f(x,z)f(z)\\
\implies p(x|z)\propto f(x,z)\\
p(y|z) &= \frac{p(y,z)}{p(z)}\\
p(y,z) \propto g(y,z)f(z)\\
\implies p(y|z)\propto g(y,z)\\
\end{align*}

> Combining these results, we see that $p(x, y | z)\propto f(x,z)g(y,z)\implies p(x, y |z) \propto p(x|z)p(y|z)$. Thus, x and y are conditionally independent. 


## Part 5

PH 3.1


### Part a

Assume Y1, . . . , Y100 are, conditional on θ, i.i.d. binary random variables with expectation θ. Write down the joint distribution of Pr(Y1 = y1, . . . , Y100 = y100|θ) in a compact form. Also write down the form of $Pr(\sum PYi = y|θ)$.


\begin{align*}
f(y_1, ..., y_{100}|\theta) &= \prod_{i=1}^{100}f(y_i|\theta)\\
&= \theta^{\sum_{i = 1}^{100}y_i}(1 - \theta)^{n - \sum_{i = 1}^{100}y_i}\\
P\left(\sum_{i = 1}^{100}Y = y_i|\theta\right) &= {n \choose \sum y_i} \theta^{\sum y_i}(1 - \theta)^{n - \sum y_i}
\end{align*}



### Part b 

For the moment, suppose you believed that θ ∈ {0.0, 0.1, . . . , 0.9, 1.0}. Given that the results of the survey were $\sum Yi = 57$, compute $Pr(\sum Yi = 57|θ)$ for each of these 11 values of θ and plot these probabilities as a function of θ.


```{r warning=FALSE, message=FALSE}
###################################### R ######################################
library(ggplot2)

# Define a function that will spit out the probability we're interested in
# computing
p_y_theta <- function(theta, n, y) {
    # y: a sum of bernouli trials
    # theta: a vector of values between 0 and 1 (can be length 1)
    # n: number of bernouli trials
    return(choose(n, y) * theta**y * (1 - theta)**(n - y))
}

# Define our primary input values (theta) and other given values (n, y)
theta <- seq(0, 1, length.out = 11)
n <- 100
y <- 57

# Figure out where the highest probability lies
max_prob_theta <- theta[
    which(p_y_theta(theta, n, y) == max(p_y_theta(theta, n, y)))
]
# Plot the outputs as nice and pretty as possible
plot_data <- data.frame(theta = theta, y = p_y_theta(theta, n, y))
ggplot(plot_data, aes(x = theta, y = y)) +
    geom_point(col = "skyblue", size = 2) +
    geom_line(col = "skyblue", lwd = 0.5, lty = 2) +
    geom_vline(xintercept = max_prob_theta, col = "firebrick", lwd = 1, lty = 2) +
    labs(title = "Pr(sum(Y) = 57 | theta)", x = "Theta", y = "Probability") +
    theme_bw() +
    theme(
        text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
```

> The plot shows a maximum at `r max_prob_theta`. 

### Part c 

Now suppose you originally had no prior information to believe one of these θ-values over another, and so Pr(θ = 0.0) = Pr(θ = 0.1) = · · · = Pr(θ = 0.9) = Pr(θ = 1.0). Use Bayes’ rule to compute $p(θ| \sum Yi = 57)$ for each θ-value. Make a plot of this posterior distribution as a function of θ.

> I want to note that I used generative AI to assist me with the following code for this problem. I derived the the math to implement, and ChatGPT assisted with the coding: 

```{r}
###################################### R ######################################

# Define the function to compute posterior probability using Bayes' theorem
p_theta_bayes <- function(theta, n, y) {
    # Compute P(Y = 57 | θ) for each θ
    likelihood <- choose(n, y) * theta^y * (1 - theta)^(n - y)

    # Prior probabilities (uniform over 11 values)
    prior <- rep(1 / length(theta), length(theta))

    # Compute P(Y = 57) as the sum of likelihood * prior over all θ
    p_y <- sum(likelihood * prior)

    # Compute posterior using Bayes' rule
    posterior <- (likelihood * prior) / p_y

    return(posterior)
}

# Define theta values and known values of n and y
theta <- seq(0, 1, length.out = 11)
n <- 100
y <- 57

# Compute the posterior probabilities
posterior_probs <- p_theta_bayes(theta, n, y)
max_prob_theta <- theta[
    which(posterior_probs == max(posterior_probs))
]

# Plot the posterior distribution
plot_data <- data.frame(theta = theta, posterior = posterior_probs)

ggplot(plot_data, aes(x = theta, y = posterior)) +
    geom_point(col = "skyblue", size = 2) +
    geom_line(col = "skyblue", lwd = 0.5, lty = 2) +
    geom_vline(col = "firebrick", lwd = 0.75, lty = 2, xintercept = max_prob_theta) +
    labs(
        title = "Posterior Probability Distribution P(θ | Y = 57)",
        x = "Theta", y = "Posterior Probability"
    ) +
    theme_bw() +
    theme(
        text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
```

> Max at $\theta$ = `r max_prob_theta`. 


### Part d 

Now suppose you allow θ to be any value in the interval [0, 1]. Using
the uniform prior density for θ, so that p(θ) = 1, plot the posterior
density $p(θ) × Pr(\sum Yi = 57|θ)$ as a function of θ.

```{r}
###################################### R ######################################
library(ggplot2)

# Define function to compute posterior density
posterior_density <- function(theta_vals, n, y) {
    # Compute likelihood P(Y = 57 | θ)
    likelihood <- choose(n, y) * theta_vals^y * (1 - theta_vals)^(n - y)

    # Since prior is uniform (1), posterior is proportional to likelihood
    posterior <- likelihood

    return(posterior)
}

# Define a continuous range of theta values
theta_vals <- seq(0, 1, length.out = 1000) # More points for smooth curve
n <- 100
y <- 57

# Compute posterior density
posterior_vals <- posterior_density(theta_vals, n, y)
max_prob_theta <- theta_vals[
    which(posterior_vals == max(posterior_vals))
]

# Normalize for better visualization (optional, but makes y-axis interpretable)
posterior_vals <- posterior_vals / max(posterior_vals)

# Plot the posterior density
plot_data <- data.frame(theta = theta_vals, posterior = posterior_vals)

ggplot(plot_data, aes(x = theta, y = posterior)) +
    geom_line(col = "skyblue", lwd = 1) +
    geom_vline(col = "firebrick", lwd = 0.75, lty = 2, xintercept = max_prob_theta) +
    labs(
        title = "Posterior Density p(θ | Y = 57)",
        x = "Theta", y = "Density (unnormalized)"
    ) +
    theme_bw() +
    theme(
        text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
```

> Max at $\theta$ = `r max_prob_theta`. 


### Part e 


```{r}
###################################### R ######################################

# Define a function to return a posterior Beta(1+57, 1+100-57) density given
# an input of theta values
posterior_beta_density <- function(theta_vals) {
    return(dbeta(theta_vals, 1 + 57, 1 + 100 - 57))
}

# Get a set of density values like we did in part d
posterior_probs <- posterior_beta_density(theta_vals)
max_prob_theta <- theta_vals[
    which(posterior_probs == max(posterior_probs))
]

# Combine into a dataframe
plot_data <- data.frame(theta = theta_vals, density = posterior_probs)

# Now let's get back to plotting!
ggplot(plot_data, aes(x = theta, y = density)) +
    geom_line(col = "skyblue", lwd = 1) +
    geom_vline(col = "firebrick", lwd = 0.75, lty = 2, xintercept = max_prob_theta) +
    labs(
        title = "Posterior Density Beta(1+57, 100+1-57)",
        x = "Theta", y = "Density"
    ) +
    theme_bw() +
    theme(
        text = element_text(family = "serif"),
        plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
    )
```

> Max at $\theta$ = `r max_prob_theta`. 


> I'm honestly pretty surprised that they have all given the same (roughly) optimal values of $\theta$. My hunch is that if I had made the first two plots as granular as the second two, I would see this same $\theta$ value of 0.569. We've used four different methods and seem to get the same results consistently. 


## Part 6 


PH 3.2


Sensitivity analysis: It is sometimes useful to express the parameters a and b in a beta distribution in terms of θ0 = a/(a + b) and n0 = a + b, so that a = θ0n0 and b = (1 − θ0)n0. Reconsidering the sample survey
data in Exercise 3.1, for each combination of θ0 ∈ {0.1, 0.2, . . . , 0.9} and n0 ∈ {1, 2, 8, 16, 32} find the corresponding a, b values and compute Pr(θ > 0.5| $\sum$ Yi = 57) using a beta(a, b) prior distribution for θ. Display the results with a contour plot, and discuss how the plot could be used to explain to someone whether or not they should believe that θ > 0.5, based on the data that $\sum$ Yi = 57.



```{r}
############################# R ########################################

library(ggplot2)
library(dplyr)

# Define prior values
theta0_values <- seq(0.1, 0.9, by = 0.1)
n0_values <- c(1, 2, 8, 16, 32)

# Create an empty dataframe to store results
results <- expand.grid(theta0 = theta0_values, n0 = n0_values)

# Observed data
y_sum <- 57  
n <- 100     

# Compute probability P(θ > 0.5) for each combination
results <- results %>%
  rowwise() %>%
  mutate(
    a = theta0 * n0,
    b = (1 - theta0) * n0,
    a_post = a + y_sum,
    b_post = b + (n - y_sum),
    prob_theta_greater_05 = 1 - pbeta(0.5, a_post, b_post)
  )

# Plot using ggplot2
ggplot(results, aes(x = theta0, y = n0, z = prob_theta_greater_05)) +
  geom_contour_filled() +
  labs(
    title = "Contour Plot of P(θ > 0.5) Given Prior (theta0, n0)",
    x = "theta0",
    y = "n0"
  ) +
  theme_minimal() +
    theme(
      text = element_text(family = "serif"),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 14)
  )

```


> The contour plot shows the probability that $\theta$ is greater than 0.5 given the data. With higher values of $\theta_0$, we see that our posterior probabilities are also higher. However, with lower values of $\theta_0$, $n_0$, which represents our prior certainty, is inversely related. That is, if I'm much more certain that $\theta_0 < 0.5$, then my posterior probability that $\theta > 0.5$ is lower. This should be intuitive and matches what we see on the plot. 




